\documentclass[10pt,a4paper]{article}

\begin{document}
\title{Provisional Goals}\maketitle{}

\section{Courses}
\label{sec:courses}
My best options are semester 2 courses, because the semester 1 courses coincide with the busy end of the provisional year. I'm only considering the courses listed on the courses pages; there are plenty more in the calendar that aren't listed as available this year. Importantly, the goal (and the continuance of my scholarship) is only met through a B+ grade in a course.

\subsection{Statistics}
\label{sec:statistics}

I have already taken the most relevant courses (\textbf{STATS 769} Data Science Practice, \textbf{STATS 782} Statistical Computing, and \textbf{STATS 760} Modern Applied Statistics)

\begin{description}
\item[STATS 730 Statistical Inference] STATS 730 gives you general-purpose skills that are required by many employers of statistical graduates. It will enable you to model real data, using likelihood-based statistical inference under the frequentist paradigm. It provides the tools and skills used by many other graduate courses on offer in this department, and it gives exposure to statistical programming in both R and the advanced optimizer TMB.\\
  \textit{This may be relevant to the ``Statistical Modelling'' part of the PhD project title}
\item[STATS 763 Advanced Regression Methodology] Semester one. 
  Generalised linear models, generalised additive models, survival analysis. Smoothing and semiparametric regression. Marginal and conditional models for correlated data. Model selection for prediction and for control of confounding. Model criticism and testing. Computational methods for model fitting, including Bayesian approaches.\\
  \textit{This may also have some relevance, in a similar sense to STATS 730}
\item[STATS 785 Topics in Statistical Data Management] One of the key purposes of STATS 785 is to introduce you to the SAS software for the purposes of statistical inference, programming and modeling.\\
  \textit{The title was promising, but I don't see a SAS programming course to be relevant}
\end{description}

\subsection{Computer Science}
\label{sec:computer-science}

The courses here actually seem more directly relevant, however the limiting factor is that I don't meet all of the prerequisites; I graduated with a Logic and Computation degree, not Computer Science, as I favoured the logical aspects of it. The extent of my CS courses are a stage 3 AI course (367), and two stage 2 CS theory courses: Algorithms (220) and Discrete Math (225). They may waive additional prerequisites, given my experience with software development in work and research, or they may not. A common pattern with the prerequisites is 230 (S1/S2) \(\to\) 335 (S2) \(\to\) 7XX.

\begin{description}
\item[COMPSCI 711 Parallel and Distributed Computing] Prerequisite: Approval of the Academic Head or nominee. Recommended Preparation: COMPSCI 335. Semester 2. Computer architectures and languages for exploring parallelism, conceptual models of parallelism, principles for programming in a parallel environment, different models to achieve interprocess communication, concurrency control, distributed algorithms and fault tolerance.\\
  \textit{Seems to really go into the details of parallelism, not sure if this lines up with what we'll be doing. Seems very relevant, however}
\item[COMPSCI 732 Software Tools and Techniques] Recommended preparation: COMPSCI 335. Semester 1. An advanced course examining research issues related to tools and techniques for software design and development. Typical topics could include advanced software development methodologies, software architectures for developing software tools, collaborative work and software engineering, open source vs closed source development, issues in advanced database systems.\\
  \textit{I have the basics (and more) understood for all of these topics, with the relevance coming from the project being the development of software.}
\item[COMPSCI 751 Advanced Topics in Database Systems] Recommended preparation: COMPSCI 220, 225. Semester 1. Database principles. Relational model, relational algebra, relational calculus, SQL, SQL and programming languages, entity-relationship model, normalisation, query processing and query optimisation, ACID transactions, transaction isolation levels, database recovery, database security, databases and XML. Research frontiers in database systems.\\
  \textit{Databases, not exactly what I'm doing, but included because I actually do meet the prerequisites, and is listed as recommended for COMPSCI 751.}
\item[COMPSCI 752 Big Data Management] Prerequisite: Approval of the Academic Head or nominee. Recommended preparation: COMPSCI 351 or equivalent. Semester 1. Big data modelling and management in distributed and heterogeneous environments. Sample topics include: representation languages for data exchange and integration (XML and RDF), languages for describing the semantics of big data (DTDs, XML Schema, RDF Schema, OWL, description logics), query languages for big data (XPath, XQuery, SPARQL), data integration (Mediation via global-as-view and local-as-vie), large-scale search (keyword queries, inverted index, PageRank) and distributed computing (Hadoop, MapReduce, Pig), big data and blockchain technology (SPARK, cryptocurrency).\\
  \textit{The description has the words Hadoop, Spark, and Big Data. The rest seems a lot less relevant, but may be useful. I get the feeling that the Hadoop and Spark material will have no more than 2 weeks dedicated to them, which won't be particularly useful to me.}
\end{description}

\section{Project-Related Goals}
\label{sec:proj-relat-goals}

Rosemary Barraclough advised me that while it is more common to have a course as a goal, it is acceptable to do 2 project-related goals. An example she gave was ``An intermediate level of experience in some software''. This will prove to be more difficult to measure, but it has the additional benefit of customisation to something relevant. Also, less chance of losing my scholarship!!

Some ideas:
\begin{itemize}
\item (From Paul) complete the literature review chapter entirely by end of first year
\item Tests on software used, such as coursera, or cloudera certification
\item SICP, metaprogramming, quotation etc.?
\item \dots{}
\end{itemize}
\end{document}
