\documentclass[a4paper,10pt]{article}

\usepackage{doc/header}

\begin{document}
\title{A Review of foreach}
\author{Jason Cairns}
\year=2020 \month=4 \day=2
\maketitle{}

\section{Introduction}
\label{sec:introduction}

foreach introduces itself on CRAN with the following description:
\begin{displaycquote}{microsoft20}
	Support for the foreach looping construct. Foreach is an idiom that
	allows for iterating over elements in a collection, without the use
	of an explicit loop counter. This package in particular is intended
	to be used for its return value, rather than for its side effects.
	In that sense, it is similar to the standard lapply function, but
	doesn't require the evaluation of a function. Using foreach without
	side effects also facilitates executing the loop in parallel.
\end{displaycquote}

From the user end, the package is conceptually simple, revolving
entirely around a looping construct and the one-off backend
registration.

The principal goal of the package, which it hasn't strayed from, is
the enabling of parallelisation through backend transparency within
the foreach construct. Notably, more complex functionality, such as
side effects and parallel recurrance, are not part of the package's
intention.

Thus, the primary driver for the practicality of the package, beyond
the support offered for parallel backends, is the backends themselves,
currently enabling a broad variety of parallel systems.

foreach is developed by Steve Weston and Hoong Ooi.

\section{Usage}
\label{sec:usage}

foreach doesn't require setup for simple serial execution, but
parallel backends require registration by the user, typically with a
single function as in the registration for doParallel,
\texttt{registerDoParallel()}.

The syntax of foreach consists of a \texttt{foreach()} function call
next to a \texttt{\%do\%} operator, and some expression to the
right\cite{weston19:_using}. Without loss in generality, the syntactic
form is given in Listing~\ref{src:syntax}.

\begin{listing}
\begin{minted}{r}
foreach(i=1:n) %do% {expr}
\end{minted}
\caption{Standard foreach syntax}\label{src:syntax}
\end{listing}

The \texttt{foreach()} function can take other arguments including
changing the means of combination along iterations, whether iterations
should be performed in order, as well as the export of environmental
variables and packages to each iteration instance.

In addition to \texttt{\%do\%}, other binary operators can be appended
or substituted. Parallel iteration is performed by simply replacing
\texttt{\%do\%} with \texttt{\%dopar\%}. Nested loops can be created
by inserting \texttt{\%:\%} between main and nested foreach functions,
prior to the \texttt{\%do\%} call\cite{weston19:_nestin_loops}. The
last step to composition of foreach as capable of list comprehension
is the filtering function \texttt{\%when\%}, which filters iterables
based on some predicate to control evaluation.


\section{Implementation}
\label{sec:implementation}

The mechanism of action in foreach is often forgotten in the face of
the atypical form of the standard syntax. Going one-by-one, the
\texttt{foreach()} function returns an iterable object,
\texttt{\%do\%} and derivatives are binary functions operating on the
iterable object returned by \texttt{foreach()} on the left, and the
expression on the right; the rightmost expression is simply captured
as such in \texttt{\%do\%}. Thus, the main beast of burder is the
\texttt{\%do\%} function, where the evaluation of the iteration takes
place.

In greater detail, \texttt{\%do\%} captures and creates environments, enabling
sequential evaluation. \texttt{\%dopar\%} captures the environment of an
expression, as well taking as a formal parameter a vector of names of libraries
used in the expression, then passing that to the backend, which will in turn do
additional work on capturing references to variables in expressions and adding
them to evaluation environment, as well as ensure packages are loaded on worker
nodes.

\texttt{\%do\%} and \texttt{\%dopar\%}, after correct error checking,
send calls to \texttt{getDoSeq()} and \texttt{getDoPar()}
respectively, which return lists determined by the registered backend,
which contain a function used backend, used to operate on the main
expression along with other environmental data.

foreach depends strongly upon the iterators package, which gives the
ability to construct custom iterators. These custom iterators can be
used in turn with the \texttt{foreach()} function, as the interface to
them is transparent.

\section{Form of Iteration}\label{sec:form-iter}

The name of the package and function interface refer to the \texttt{foreach}
programming language construct, present in many other languages.
By definition, the \texttt{foreach} construct performs traversal over some
collection, not necessarily requiring any traversal order.
In this case, the collection is an iterator object or an object coercible to
one, but in other languages with foreach as part of the core language, such as
python (whose for loop is actually only a foreach loop), collections can
include sets, lists, and a variety of other classes which have an
\texttt{\_\_iter\_\_} and \texttt{\_\_next\_\_} defined\cite{python2020iter}.

Due to the constraints imposed by a foreach construct, loop optimisation is
simplified relative to a for loop, and the lack of explicit traversal ordering
permits parallelisation, which is the primary reason for usage of the
\texttt{foreach} package.
The constraints are not insignificant however, and they do impose a limit on
what can be expressed through their usage.
Most notably, iterated functions, wherein the function depends on it's prior
output, are not necessarily supported, and certainly not supported in parallel.
This is a result of the order of traversal being undefined, and when order is
essential to maintain coherent state, as in iterated functions, the two
concepts are mutually exclusive.

In spite of the constraints, iterated functions can actually be emulated in
foreach through the use of destructive reassignment within the passed
expression, or through the use of stateful iterators.
Examples of both are given in listings \ref{src:serial} and \ref{src:serial-iter}.

\begin{listing}
\begin{minted}{R}
x <- 10
foreach(i=1:5) %do% {x <- x+1}
\end{minted}
\caption{Serial iterated function through destructive reassignment}\label{src:serial}
\end{listing}

\begin{listing}
\begin{minted}{R}
addsone <- function(start, to) {
	nextEl <- function(){
		start <<- start + 1
		if (start >= to) {
			stop('StopIteration')
		}
		start}
	obj <- list(nextElem=nextEl)
	class(obj) <- c('addsone', 'abstractiter', 'iter')
	obj
}

it <- addsone(10, 15)
nextElem(it)

foreach(i = addsone(10, 15), .combine = c) %do% i
\end{minted}
\caption{Serial iterated function through creation of a stateful iterator}\label{src:serial-iter}
\end{listing}

As alluded to earlier, the functionality breaks down when attempting to run
them in parallel.
Listings \ref{src:parallel} and \ref{src:parallel-iter} demonstrate attempts to
evaluate these iterated functions in parallel.
They only return a list of 5 repetitions of the same ``next'' number, not
iterating beyond it.

\begin{listing}
\begin{minted}{R}
cl <- makeCluster(2)
doParallel::registerDoParallel(cl)
x <- 10
foreach(i=1:5) %dopar% {x <- x+1}
\end{minted}
\caption{Parallel Iteration attempt through destructive reassignment}\label{src:parallel}
\end{listing}

\begin{listing}
\begin{minted}{R}
doParallel::registerDoParallel
foreach(i = addsone(10, 15), .combine = c) %dopar% i
\end{minted}
\caption{Parallel Iteration attempt through a stateful iterator}\label{src:parallel-iter}
\end{listing}

\section{Extensions}
\label{sec:extensions}

The key point of success in foreach is it's backend extensibility,
without which, foreach would lack any major advantages over a standard
\texttt{for} loop.

Other parallel backends are enabled through specific functions made
available by the foreach package. The packages define their parallel
evaluation procedures with reference to the iterator and accumulator
methods from foreach.

Numerous backends exist, most notably:
\begin{description}
	\item[doParallel] the primary parallel backend for foreach, using the
	      parallel package\cite{corporation19}.
	\item[doRedis]  provides a Redis backend, through the redux package\cite{lewis20}.
	\item[doFuture] uses the future package to make use of future's many
	      backends\cite{bengtsson20do}.
	\item[doAzureParallel] allows for direct submission of parallel
	      workloads to an Azure Virtual Machine\cite{hoang20}.
	\item[doMPI] provides MPI access as a backend, using Rmpi\cite{weston17}.
	\item[doRNG] provides for reproducible random number usage within
	      parallel iterations, using L'Ecuyer's method; provides
	      \texttt{\%dorng\%}\cite{gaujoux20}.
	\item[doSNOW] provides an ad-hoc cluster backend, using the snow
	      package\cite{dosnow19}.
\end{description}

\section{Relevance}
\label{sec:relevance}

foreach serves as an example of a well-constructed package supported
by it's transparency and extensibility.

For packages looking to provide any parallel capabilities, a foreach
extension would certainly aid it's potential usefulness and
visibility.

\printbibliography{}
\end{document}
