\documentclass[10pt,a4paper]{article}

\usepackage{doc/header}

\begin{document}
\title{A Detail of future}
\author{Jason Cairns}
\year=2020 \month=3 \day=19
\maketitle{}

\section{Overview}
\label{sec:overview}
\nocite{bengtsson19:_futur_r}

future is introduced with the following summary: \blockquote{The
purpose of this package is to provide a lightweight and unified
future API for sequential and parallel processing of R expression
via futures. The simplest way to evaluate an expression in parallel
is to use \texttt{x \%<-\% { expression }} with
\texttt{plan(multiprocess)}. This package implements sequential,
multicore, multisession, and cluster futures. With these, R
expressions can be evaluated on the local machine, in parallel a set
of local machines, or distributed on a mix of local and remote
machines. Extensions to this package implement additional backends
for processing futures via compute cluster schedulers etc. Because
of its unified API, there is no need to modify any code in order
switch from sequential on the local machine to, say, distributed
processing on a remote compute cluster. Another strength of this
package is that global variables and functions are automatically
identified and exported as needed, making it straightforward to
tweak existing code to make use of futures.\cite{bengtsson20}}
futures are abstractions for values that may be available at some
point in the future, taking the form of objects possessing state,
being either resolved and therefore available immediately, or
unresolved, wherein the process blocks until resolution.

futures find their greatest use when run asynchronously. The future
package has the inbuilt capacity to resolve futures asynchronously,
including in parallel and through a cluster, making use of the
parallel package. This typically runs a separate process for each
future, resolving separately to the current R session and modifying
the object state and value according to it's resolution status.

\section{Comparison with Substitution and Quoting}
\label{sec:comparison-with-non}

R lays open a powerful set of metaprogramming functions, which bear
similarity to future. R expressions can be captured in a
\texttt{quote()}, then evaluated in an environment with \texttt{eval()}
at some point in the future. Additionally, \texttt{substitute()}
substitutes any variables in the expression passed to it with the
values bound in an environment argument, thus allowing ``non-standard
evaluation'' in functions.

future offers a delay of evaluation as well, however such a delay is
not due to manual control of the programmer through \texttt{eval()}
functions and the like, but due to background computation of an
expression instead.

\section{Example Usage}
\label{sec:examples}

Through substitution and quoting, R can, for example, run a console within the
language. Futures allows the extension of this to a parallel evaluation scheme.
Listing \ref{src:console} gives a simple implementation of this idea: a console
that accepts basic expressions, evaluating them in the background and
presenting them upon request when complete.  Error handling and shared
variables are not implemented.

\begin{listing}
\inputminted{r}{R/review-future.R}
\caption{Usage of future to implement a basic multicore console}
\label{src:console}
\end{listing}

\section{Extension Packages}
\label{sec:extension-packages}

\begin{description}
	\item[doFuture]\cite{bengtsson20do} provides an adapter for
	      foreach\cite{microsoft20} that works on a future-based backend. Note
	      that this does does not return foreach() calls as futures. The
	      multicore features enabled with future are redundant over the
	      existing parallel package, but because future backends can include
	      other clusters, such as those provided by batchtools, there is some
	      additional functionality, including additional degrees of control
	      over backends.
	\item[future.batchtools]\cite{bengtsson19batch} provides a future API
	      for batchtools\cite{lang17}, or equivalently, a batchtools backend
	      for future. This allows the use of various cluster schedulers such
	      as TORQUE, Slurm, Docker Swarm, as well as custom cluster functions.
	\item[future.apply]\cite{bengtsson20apply} provides equivalent
	      functions to R's \texttt{apply} procedures, with a future backend
	      enabling parallel, cluster, and other functionality as enabled by
	      backends such as batchtools through future.batchtools.
	\item[future.callr]\cite{bengtsson19callr} provides a
	      callr\cite{csardi20} backend to future, with all of the associated
	      advantages and overhead. Callr \enquote{call[s] R from R}. It
	      provides functions to run expressions in a background R process,
	      beginning a new session. An advantage of callr is that it allows
	      more than 125 connections, due to not making use of R-specific
	      connections. Additionally, no ports are made use of, unlike the
	      SOCKcluster provided by the snow component of parallel.
	\item[furrr]\cite{vaughan18} allows the use of future as a backend to
	      purrr functions. purrr is a set of functional programming tools for
	      R, including map, typed map, reduce, predicates, and monads. Much of
	      it is redundant to what already exists in R, but it has the
	      advantage and goal of adhering to a consistent standard.
\end{description}

\section{Further Considerations}
\label{sec:further-considerations}

One initial drawback to future is the lack of callback functionality, which
would open enormous potential. However, this feature is made available in the
\textit{promises} package, which has been developed by Joe Cheng at RStudio,
which allows for user-defined handlers to be applied to futures upon
resolution\cite{Cheng19}.

Issues that aren't resolved by other packages include the copying of objects
referenced by future, with mutable objects thereby unable to be directly
updated by future (though this may be ameliorated with well-defined callbacks).
This also means that data movement is mandatory, and costly; future raises an
error if the data to be processed is over 500Mb, though this can be overridden.

Referencing variables automatically is a major unsung feature of future, though
it doesn't always work reliably; future relies on code inspection, and allows a
\texttt{global} parameter to have manual variable specification.

It seems likely that the future package will have some value to it's use,
especially if asynchronous processing is required on the R end; it is the
simplest means of enabling asynchrony in R without having to manipulate
networks or threads.

\printbibliography{}

\end{document}
